\section{Future Work}
Future research should investigate the long-term cognitive and behavioral impact of AI-assisted development on programmers with varying levels of expertise. While existing work primarily focuses on short-term productivity improvements, there remains limited understanding of how continuous reliance on LLMs affects learning, problem-solving ability, debugging skills, and code comprehension over time. Longitudinal studies conducted in real-world industrial environments would provide deeper insights into how developers adapt their workflows and reasoning strategies when collaborating with AI systems.

Further research can focus on improving the reliability, contextual reasoning, and scalability of LLM-assisted programming systems. Current models still struggle with complex logical reasoning, multi-file dependencies, and large-scale software architectures. Future systems should better integrate contextual awareness, verification mechanisms, and explainability features to support trustworthy collaboration in professional software engineering environments. Research is also needed to explore hybrid interaction models that combine conversational, generative, and retrieval-based assistance more effectively across different stages of the SDLC.

Another promising direction for future research involves the emergence of agentic AI workflows in software engineering. Unlike current AI-assisted programming tools that primarily respond to developer prompts, agentic systems are capable of autonomously planning, decomposing, and executing multi-step development tasks with limited human intervention. Such systems may fundamentally reshape developer responsibilities by shifting humans further toward supervision, validation, and high-level architectural decision-making. Future work should investigate how agentic workflows influence developer cognition, collaboration, trust, and accountability, particularly in large-scale or safety-critical software systems. Additionally, research is needed to understand the reliability, transparency, and coordination challenges introduced by increasingly autonomous AI-driven development environments.
